---
title: "inspect labelled cells"
author: "Daniel Schulz"
date: "2024-07-26"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Please set this path to the folder which contains all the data downloaded from Zenodo
mount_path <- "D:/Data/classification_public_for_zenodo_upload/"
```

In this script we will inspect the manually gated cells

Note: gating was performed in two large batches (V1 and V3). We will inspect the labelled cells of both batches here.

```{r load-libraries}
library(dittoSeq)
library(scran)
library(viridis)
library(scater)
```


```{r load-data}
sce <- readRDS(paste0(mount_path,"sce_labelled_V1.rds"))
sce_new <- readRDS(paste0(mount_path,"sce_labelled_V3.rds"))
```

```{r define-color-vector, fig.height=7, fig.width=7}
cell_types <- c("B", "plasma", "CD4", "CD8", "Treg", "BnT",
                "DC", "HLADR", "MacCD163", "Neutrophil", "NK", "pDC", "Mural", 
                "Tumor", "unlabelled", "undefined")

names(cell_types) <- cell_types
cell_types["B"] <- "red3"
cell_types["plasma"] <- "goldenrod2"
cell_types["CD4"] <- "palevioletred1"
cell_types["CD8"] <- "darkorchid"
cell_types["Treg"] <- "maroon4"
cell_types["BnT"] <- "darkorange4"
cell_types["DC"] <- "green4"
cell_types["HLADR"] <- "darkgreen"
cell_types["MacCD163"] <- "green"
cell_types["Neutrophil"] <- "blue1"
cell_types["NK"] <- "yellow"
cell_types["pDC"] <- "deepskyblue"
cell_types["Mural"] <- "tomato"
cell_types["Tumor"] <- "sienna4"
cell_types["unlabelled"] <- "gray"
cell_types["undefined"] <- "black"

metadata(sce)$colour_vectors$cell_types <- cell_types
metadata(sce_new)$colour_vectors$cell_types <- cell_types
```

# Heatmap V1 labelling

## markers used for labelling
```{r heatmap-V1, fig.width=20, fig.height=15}
lab_sce <- cbind(sce,sce_new)
lab_sce <- lab_sce[,lab_sce$cell_labels != "unlabelled"]
agr_sce <- aggregateAcrossCells(lab_sce, 
                                ids = colData(lab_sce)[,c("sample_id", "cell_labels")], 
                                statistics = "mean")
assay(agr_sce, "exprs") <- asinh(counts(agr_sce))
colnames(agr_sce) <- paste0(agr_sce$sample_id, "_", 
                            agr_sce$cell_labels)

# Define markers that were used for gating
cur_markers <- c("Ecad", "CarbonicAnhydrase", "HLADR", 
                 "CD11c", "CD163", "CD303", "CD15", "MPO", "CD38",
                 "SMA", "PDGFRb", "CD20", "CD7", "CD3", "CD8a", "CD4", "FOXP3")

# Non-scaled
dittoHeatmap(agr_sce[cur_markers,], assay = "exprs",
            annot.by = c("sample_id", "cell_labels", "Indication"), 
            cluster_rows = FALSE,
            scale = "none", heatmap.colors = viridis(100), order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce)$colour_vectors$cell_types))

ncol(lab_sce)
```

## all good markers
```{r heatmap-V1, fig.width=20, fig.height=15}
lab_sce <- cbind(sce,sce_new)
lab_sce <- lab_sce[,lab_sce$cell_labels != "unlabelled"]
agr_sce <- aggregateAcrossCells(lab_sce, 
                                ids = colData(lab_sce)[,c("sample_id", "cell_labels")], 
                                statistics = "mean")
assay(agr_sce, "exprs") <- asinh(counts(agr_sce))
colnames(agr_sce) <- paste0(agr_sce$sample_id, "_", 
                            agr_sce$cell_labels)

# Define markers that were used for gating
good_markers <- rownames(sce)[!grepl("DNA|Histone",rownames(sce))]

# Non-scaled
dittoHeatmap(agr_sce[good_markers,], assay = "exprs",
            annot.by = c("sample_id", "cell_labels", "Indication"), 
            cluster_rows = FALSE,
            scale = "none", heatmap.colors = viridis(100), order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce)$colour_vectors$cell_types))

# Centered and scaled
png(filename = "C:/Users/dschulz/switchdrive/IMMUcan_new/Manuscripts/Workflow2022/Figures/Figure_6/gated_cells_heatmap.png",
    width = 1000,
    height = 1000,
    units = "px")
dittoHeatmap(agr_sce[good_markers,], assay = "exprs",
            annot.by = c("cell_labels", "Indication"), 
            cluster_rows = TRUE, order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce)$colour_vectors$cell_types),
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-3,3, length.out = 101))
dev.off()

pdf(file =  "C:/Users/dschulz/switchdrive/IMMUcan_new/Manuscripts/Workflow2022/Figures/Figure_6/gated_cells_heatmap.pdf",
    width = 8,
    height = 6)
dittoHeatmap(agr_sce[good_markers,], assay = "exprs",
            annot.by = c("cell_labels", "Indication"), 
            cluster_rows = TRUE, order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce)$colour_vectors$cell_types),
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-3,3, length.out = 101))
dev.off()
```

## Batch correction
```{r}
library(batchelor)
library(BiocParallel)
rowData(sce)$use_channel <- !grepl("DNA|Histone", rownames(sce))

out <- fastMNN(sce, batch = sce$sample_id,
               subset.row = rowData(sce)$use_channel,
               assay.type = "exprs", 
               BPPARAM = MulticoreParam(workers = 16, RNGseed = 221101))

# Transfer the correction results to the main spe object
reducedDim(sce, "fastMNN") <- reducedDim(out, "corrected")
```

## plot UMAP
```{r}
set.seed(221101)
sce <- runUMAP(sce,
               subset_row = good_markers,
               name = "UMAP",
               assay.type = "exprs" )

cur_unlab_sce <- sce[,sce$cell_labels == "unlabelled"]
cur_lab_sce <- sce[,sce$cell_labels != "unlabelled"]

ggplot() +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), 
              data = data.frame(UMAP_1 = reducedDim(cur_unlab_sce, "UMAP")[,1],
                                UMAP_2 = reducedDim(cur_unlab_sce, "UMAP")[,2],
                                cell_labels = colData(cur_unlab_sce)$cell_labels)) +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), size = 0.5, 
            data = data.frame(UMAP_1 = reducedDim(cur_lab_sce, "UMAP")[,1],
                              UMAP_2 = reducedDim(cur_lab_sce, "UMAP")[,2],
                              cell_labels = colData(cur_lab_sce)$cell_labels)) +
    scale_color_manual(values = metadata(sce)$colour_vectors$cell_types) + 
    theme_bw()
```

## plot batch corrected UMAP
```{r}
set.seed(221101)
sce <- runUMAP(sce, dimred = "fastMNN", name = "UMAP_fastMNN")

cur_unlab_sce <- sce[,sce$cell_labels == "unlabelled"]
cur_lab_sce <- sce[,sce$cell_labels != "unlabelled"]

ggplot() +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), 
              data = data.frame(UMAP_1 = reducedDim(cur_unlab_sce, "UMAP_fastMNN")[,1],
                                UMAP_2 = reducedDim(cur_unlab_sce, "UMAP_fastMNN")[,2],
                                cell_labels = colData(cur_unlab_sce)$cell_labels)) +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), size = 0.5, 
            data = data.frame(UMAP_1 = reducedDim(cur_lab_sce, "UMAP_fastMNN")[,1],
                              UMAP_2 = reducedDim(cur_lab_sce, "UMAP_fastMNN")[,2],
                              cell_labels = colData(cur_lab_sce)$cell_labels)) +
    scale_color_manual(values = metadata(sce)$colour_vectors$cell_types) + 
    theme_bw()
```


# Heatmap V3 labelling

## markers used for labelling
```{r heatmap-V1, fig.width=20, fig.height=15}
lab_sce <- sce_new[,sce_new$cell_labels != "unlabelled"]
agr_sce_new <- aggregateAcrossCells(lab_sce_new, 
                                ids = colData(lab_sce_new)[,c("sample_id", "cell_labels")], 
                                statistics = "mean")
assay(agr_sce_new, "exprs") <- asinh(counts(agr_sce_new))
colnames(agr_sce_new) <- paste0(agr_sce_new$sample_id, "_", 
                            agr_sce_new$cell_labels)

# Define markers that were used for gating
cur_markers <- c("Ecad", "CarbonicAnhydrase", "Ki67", "CD14", "HLADR", 
                 "CD11c", "CD163", "CD303", "CD68", "CD15", "MPO", "CD38",
                 "SMA", "PDGFRb", "CD20", "CD7", "CD3", "CD8a", "CD4", "GrzB", "FOXP3", "PD1")

# Non-scaled
dittoHeatmap(agr_sce_new[cur_markers,], assay = "exprs",
            annot.by = c("sample_id", "cell_labels", "Indication"), 
            cluster_rows = FALSE,
            scale = "none", heatmap.colors = viridis(100), order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce_new)$colour_vectors$cell_types))

# Centered and scaled
dittoHeatmap(agr_sce_new[cur_markers,], assay = "exprs",
            annot.by = c("sample_id", "cell_labels", "Indication"), 
            cluster_rows = FALSE, order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce_new)$colour_vectors$cell_types),
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-3, 3, length.out = 101))
```

## all good markers
```{r heatmap-V1, fig.width=20, fig.height=15}
lab_sce_new <- sce_new[,sce_new$cell_labels != "unlabelled"]
agr_sce_new <- aggregateAcrossCells(lab_sce_new, 
                                ids = colData(lab_sce_new)[,c("sample_id", "cell_labels")], 
                                statistics = "mean")
assay(agr_sce_new, "exprs") <- asinh(counts(agr_sce_new))
colnames(agr_sce_new) <- paste0(agr_sce_new$sample_id, "_", 
                            agr_sce_new$cell_labels)

# Define markers that were used for gating
good_markers <- rownames(sce_new)[!grepl("DNA|Histone",rownames(sce_new))]

# Non-scaled
dittoHeatmap(agr_sce_new[good_markers,], assay = "exprs",
            annot.by = c("sample_id", "cell_labels", "Indication"), 
            cluster_rows = FALSE,
            scale = "none", heatmap.colors = viridis(100), order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce_new)$colour_vectors$cell_types))

# Centered and scaled
dittoHeatmap(agr_sce_new[good_markers,], assay = "exprs",
            annot.by = c("sample_id", "cell_labels", "Indication"), 
            cluster_rows = FALSE, order.by = "cell_labels",
            annotation_colors = list(cell_labels = metadata(sce_new)$colour_vectors$cell_types),
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-3, 3, length.out = 101))
```

## Batch correction
```{r}
library(batchelor)
library(BiocParallel)
rowData(sce_new)$use_channel <- !grepl("DNA|Histone", rownames(sce_new))

out <- fastMNN(sce_new, batch = sce_new$sample_id,
               subset.row = rowData(sce_new)$use_channel,
               assay.type = "exprs", 
               BPPARAM = MulticoreParam(workers = 16, RNGseed = 221101))

# Transfer the correction results to the main spe object
reducedDim(sce_new, "fastMNN") <- reducedDim(out, "corrected")
```

## plot UMAP
```{r}
set.seed(221101)
sce_new <- runUMAP(sce_new,
               subset_row = good_markers,
               name = "UMAP",
               assay.type = "exprs" )

cur_unlab_sce_new <- sce_new[,sce_new$cell_labels == "unlabelled"]
cur_lab_sce_new <- sce_new[,sce_new$cell_labels != "unlabelled"]

ggplot() +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), 
              data = data.frame(UMAP_1 = reducedDim(cur_unlab_sce_new, "UMAP")[,1],
                                UMAP_2 = reducedDim(cur_unlab_sce_new, "UMAP")[,2],
                                cell_labels = colData(cur_unlab_sce_new)$cell_labels)) +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), size = 0.5, 
            data = data.frame(UMAP_1 = reducedDim(cur_lab_sce_new, "UMAP")[,1],
                              UMAP_2 = reducedDim(cur_lab_sce_new, "UMAP")[,2],
                              cell_labels = colData(cur_lab_sce_new)$cell_labels)) +
    scale_color_manual(values = metadata(sce_new)$colour_vectors$cell_types) + 
    theme_bw()
```

## plot batch corrected UMAP
```{r}
set.seed(221101)
sce_new <- runUMAP(sce_new, dimred = "fastMNN", name = "UMAP_fastMNN")

cur_unlab_sce_new <- sce_new[,sce_new$cell_labels == "unlabelled"]
cur_lab_sce_new <- sce_new[,sce_new$cell_labels != "unlabelled"]

ggplot() +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), 
              data = data.frame(UMAP_1 = reducedDim(cur_unlab_sce_new, "UMAP_fastMNN")[,1],
                                UMAP_2 = reducedDim(cur_unlab_sce_new, "UMAP_fastMNN")[,2],
                                cell_labels = colData(cur_unlab_sce_new)$cell_labels)) +
    geom_point(aes(x = UMAP_1, y = UMAP_2, colour = cell_labels), size = 0.5, 
            data = data.frame(UMAP_1 = reducedDim(cur_lab_sce_new, "UMAP_fastMNN")[,1],
                              UMAP_2 = reducedDim(cur_lab_sce_new, "UMAP_fastMNN")[,2],
                              cell_labels = colData(cur_lab_sce_new)$cell_labels)) +
    scale_color_manual(values = metadata(sce_new)$colour_vectors$cell_types) + 
    theme_bw()
```
